Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-02 aistats/paper.tex
Start adding first few sections to paper

\documentclass{article}

\usepackage[top=2cm, bottom=2cm]{geometry}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}           % define for loops

@@

\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsection*{\bibname}}

@@

% frames with colored background
\definecolor{light-gray}{gray}{0.85}
\definecolor{myblue}{RGB}{199, 235, 255}
\usepackage[framemethod=TikZ]{mdframed} 
\usetikzlibrary{shadows}
% Using the pgf-blur package instead 
% https://ctan.math.illinois.edu/graphics/pgf/contrib/pgf-blur/pgf-blur.pdf
\usetikzlibrary{shadows.blur}

\mdfdefinestyle{blurry-shadow}{
	linewidth=0pt,
    roundcorner=0pt,
	shadow=true,
 	apptotikzsetting={\tikzset{mdfshadow/.style={blur shadow={%
 		shadow xshift=1pt,
 		shadow yshift=-1pt,
 		shadow scale=1.00,
 		shadow opacity = 50,
 		%shadow blur extra rounding,
 		shadow blur steps = 10,
 		shadow blur radius = 5pt,
 		every shadow
 	}}}},
	linewidth=0pt,
 }

\newmdenv[% box with grey background
	style=blurry-shadow,
    backgroundcolor=gray!15,
]{example}
\newmdenv[
	style=blurry-shadow,
	backgroundcolor=Red!30,
]{important}

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

@@

\title{Convergence rate of MAP estimates \\
for the exponential family}
\author{R\'emi Le Priol}
\date{October 2020}

%% Faster processing
%\renewcommand\appendix{\end{document}}

@@

\maketitle

\section{Background}

\paragraph{Motivation.} 
We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.
We want the simplest one.
We hope  to get a new result by combining tools from statistics and optimization. 
\paragraph{Exponential Family.}
The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model 
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
where $\logpart$ is the log-partition function (aka normalization factor) 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}

\begin{example}
	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
	The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
\end{align}
\end{example}

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[ht]
	\centering
	\includegraphics{figs/duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$. Figure reproduced from \citet{kunstner2020homeomorphic}.}
	\label{fig:duality}
\end{figure}
\begin{example}
For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
\end{example}

\paragraph{Conjugate Prior}
\TODO{Under-parametrized conjugate prior ?}
One conjugate prior for $p(X|\natp)$ is
\begin{align}
    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
\end{align}
where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp ; \natp_0)
    & = \logpart(\natp) - \logpart(\natp_0) 
    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
\end{align}
with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
We can re-write this prior as 
\begin{align}
    p(\natp) \propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \; ,
\end{align}
which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.

\begin{example}
In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
In particular,
\begin{align}
	p(\natp) 
	&\propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \\
    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$. After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
\end{example}

\paragraph{Maximum A Posteriori (MAP).}
The negative log-likelihood of the prior is
\begin{align*}
    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
\begin{align}
    -\log p(\cD|\natp)p(\natp) 
    = (n_0+n) \logpart (\natp) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
\end{align}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
\begin{align}
	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
\end{align}
The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.

\section{Open Problem}
For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{align}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}
\begin{important}
The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?
\end{important}

@@

Below we review all the attempts we made on this problem.

\tableofcontents

\section{Gaussian Variance Example}

\begin{example}
	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}
\end{example}
In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{figs/bregmandef.pdf}
	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.54\textwidth]{figs/asymptote.pdf}
	\includegraphics[width=.43\textwidth]{figs/fewsamples.pdf}
	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$. Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.  
	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
	\textbf{Right:} the first few samples significantly deviate from this behavior. In fact, for $n=1$ and $n=2$, the expected value is infinite, but we have a closed form solution and a simple upper-bound for $n>2$.
	}
	\label{fig:curves}
\end{figure}

\begin{example}
	\begin{theorem}[MLE Tight Bound]
	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
\end{example}
This upper bound is asymptotically tight.
We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
 There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}. The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension. We do not perform these calculations for now as we want to focus on the simplest possible examples.

As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
We now present a lemma on its expected inverse.
\begin{lemma}[Expected MAP Natural Parameter]
	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
	\begin{equation}
		\frac{\mu^*}{\mu_n}
		\leq \expect{\frac{\mu^*}{\MAPm}} 
		= \expect{\frac{\MAPt}{\natp^*}} 
		\leq \frac{n_0 +n}{a+ (n-2)_+}
		\label{eq:theta_expectation_bound}
	\end{equation}
	where $(x)_+ = \max(0,x)$.
\end{lemma}
Note that the lower bound is a simple consequence of the convexity of the inverse function. 
Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
\begin{example}
\begin{theorem}[MAP Bound]
 Let us  define
 \begin{align}
	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
 \begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}
\end{equation}
\end{theorem}
\end{example}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$. We are basically losing a factor 2 compared to $1/2n$.

Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
\begin{example}
\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
\begin{equation}
	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
	\leq \inv{n-2} \; .
	\label{eq:MLE_loose_rate}
\end{equation}
\end{corollary}
\end{example}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
	\caption{
	Convergence rate of MAP~\eqref{eq:MAP_rate} 
	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
	We report these curves for various values of $n$, and $n_0=5$. 
	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$). 
	This observation hints towards the optimality of MLE for large $n$. 
	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
	}
\end{figure}

\TODO{add equations for other families, eg gamma or Dirichlet, or Categorical..}

>>>>

\documentclass[twoside]{article}
\usepackage{aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%\usepackage[accepted]{aistats2022}

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\graphicspath{{../figs/}}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}

@@

@@

% my packages
\usepackage{../math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

@@

@@

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Looking for Convergence Rates\\ for the MAP of the Exponential Family}

\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner \And  Damien Scieur \And Simon Lacoste-Julien }

\aistatsaddress{ Mila \And  UBC \And SAIT SAIL \And Mila} 
]

\begin{abstract}
We raise the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate, or a conjugate maximum a posteriori for the exponential family. 
Surprisingly, we found no solution to this problem in the literature -- we are not able to tell how many samples we need to fit a gaussian within a few bits of the true distribution.
After displaying some properties and special cases of this problem, 
we show it is a special case of several optimization algorithms, but it falls out of their scopes, thus highlighting range of progress in the analysis of these algorithms.
\end{abstract}

\section{Plan}

\begin{enumerate}
	\item intro to exponential family, and density estimation
	\item the thing we want to bound
	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
	\item insight : bias-variance decomposition
	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
\end{enumerate}

Open questions
\begin{itemize}
	\item does a base measure change anything ?
	\item is there multiple conjugate priors ?
	\item misspecified case. Do we have a   formula ?
	\item make sure the gaussian entropy is not SC.
\end{itemize}

\section{Introduction and Background}
\RLP{Goal = disseminate related work to make it look nice.}
Exponential Families are an elegant and lean way to model a wide variety of data : binary, categorical, natural numbers, positive float, long or short tailed... 
They are literally the linear model of probabilities.
The exponential family for data $X\in\cX$  with sufficient statistic $T$ and natural parameter $\natp$ is the model 
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
where $\logpart$ is the log-partition function -- i.e. the normalization factor
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx 
\end{align}
where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
Note that an exponential family is entirely specified by its support set $\cX$ and its sufficient statistic $T$.
This simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.

\paragraph{Duality}
The logpartition function $\logpart$ verifies the two following identities
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
The second identity entails that $\logpart$ is strictly-convex. 
At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
\begin{align}
	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
\end{align}
It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
\begin{figure}[ht]
	\centering
	\includegraphics{duality}
	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$. Figure reproduced from \citet{kunstner2020homeomorphic}.}
	\label{fig:duality}
\end{figure}

\section{Open Problem}

For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{multline}
    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
\end{multline}
For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
\begin{align}
	\KL( p(.|\natp^*) ; p(.|\natp)) 
	= \bregman (\natp ; \natp^*)  \; .
\end{align}
There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
\begin{align}
	\bregman (\natp ; \natp^*)
    %= \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
    = \bregmanconj ( \meanp^* ; \meanp)
\end{align}
so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
\begin{align}
\boxed{
	\KL( p(.|\natp^*) ; p(.|\natp))
    = \bregman (\natp ; \natp^*)
    = \bregmanconj ( \meanp^* ; \meanp) \; .
}
\end{align}

The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
	\label{eq:bregmanMAP}
	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
\end{align}
where the outer expectation is on the dataset $X_1, \dots, X_n$?

@@

\section{Examples}
\subsection{Gaussian Mean}

\subsection{Gaussian Variance}
The trailing example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
The density of a centered normal variable is
\begin{align}
	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
\end{align}
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
Mean and natural parameters are roughly inverse of each other
\begin{align}
	\natp = -\inv{2 \mu}\; .
\end{align}
Now we can match the log-likelihood with the exponential family template to get the log-partition function.
\begin{align}
	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
	= x^2 \natp - \logpart(\natp) \\
	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
\end{align}

We can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
\begin{align}
	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
\end{align}
We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
\begin{align}
	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
\end{align}
which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.

The logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
In particular,
\begin{align}
	p(\natp) 
	&\propto 
    \exp( -n_0 \logpart (\natp) 
    + \langle n_0 \mu_0, \natp \rangle ) \\
    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
\end{align}
from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$. After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.

Both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
\begin{align}
	\KL( \sigma_*^2 ; \sigma_n^2 ) 
	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
	\bregmanconj( \mu_*; \mu_n) 
	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
	\bregman( \natp_n; \natp_* ) 
	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
\end{align}

In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
\begin{align}
	\phi(z) := \half (z - 1 - \log(z)) \\
	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
\end{align}
as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
\begin{align}
	1 - \inv{z} \leq \log(z) 
	\label{eq:log_bound} 
	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
\end{align}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.4\textwidth]{bregmandef.pdf}
	\includegraphics[width=.4\textwidth]{phi.pdf}
	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
	\label{fig:phi}
\end{figure}

\begin{figure}[ht]
	\centering
\includegraphics[width=.48\textwidth]{asymptote.pdf}
\includegraphics[width=.48\textwidth]{fewsamples.pdf}
	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$. Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.  
	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
	\textbf{Right:} the first few samples significantly deviate from this behavior. In fact, for $n=1$ and $n=2$, the expected value is infinite, but we have a closed form solution and a simple upper-bound for $n>2$.
	}
	\label{fig:curves}
\end{figure}

\bibliographystyle{apalike}
\bibliography{../references.bib}
